---
title: "Version Control With Git"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: false
  fonttheme: "serif"
  includes:
   in_header: slides_preamble.tex
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = T, warning = FALSE, message = FALSE)
```

# What is Git?

* Version control software keeps track of all the versions or revisions to a set of files in an efficient and orderly manner. 
* Git is version control software, initially developed for version control of the Linux operating system kernel.
* It is now extremely widely used for almost all kinds of software development projects. 
* Git is characterized by the following features or goals:
  * Nonlinear development. 
  * Distributed and decentralized. 
  * Speed and efficiency.
  * Open source.

# How Git works internally (very briefly)

* Git works on whole "projects", which are simply directories, not individual files.
* Each project is completely self-contained, and completely independent of every other. 
* It keeps a "snapshot" of the state of the project as it develops. These are *commits*.
* Git copies every version of every file in every commit. Calculates its sha-1 hash. Compresses it. Names it by its hash. These are known as *blobs*. 
* It creates a hash tree (Merkle tree) from the files in each commit (the *tree object*). 
* Each commit stores the tree hash, parent commit, author info, log message, etc, and calculate it sha-1 hash.

# Distributed and decentralized

* The Git project is completely self-contained. All the meta data and history of the project is contained in `.git`.
* There is no centralized server. No synchronization. No edit locks. 
* Collaboration is via cloning and re-emerging.
* E.g. a project can be *cloned*, developed independently, and possibly re-merged. 
* Usually developers use a *remote* host with a "bare" repository, clone it, develop locally, *commit* and then *push* back to and *pull* from the remote host. The *push* and *pull* are merges.
* GitHub is one of the most widely used hosting sites (but there are others, e.g. BitBucket; and running your own git hosting server is simple and inexpensive).

# What should we (academics, researchers, scientists) care?

* Git (and a Git server) provides a means of sharing your *research compendium* and its updates, and this can be done from the beginning of the research and continually throughout its entire development.
* Sharing can be with the public or just with collaborators
* But it's not just another "dropbox": It is a decentralized version control system. 
  * It allows open, distributed, collaborative development of research and data analysis code, like the open source development model for software generally.
  * Even for small teams, it provides a means for efficient and organized development of source code, including (and especially) for writing the source code for our articles, slides, etc (no more of this: http://phdcomics.com/comics/archive_print.php?comicid=1531). 
  * Even for individuals, it allows for efficient and organized development.
  * And it is (almost) endlessly fault tolerant. So long as there is one clone somewhere, the entire project and its version history is preserved entirely.

# Git: Tiny tutorial

- Start by cloning a remote repository:
```{r, engine = 'bash', eval = FALSE}
      git clone https://github.com/yihui/knitr.git
      cd knitr
      git log # Read all the commit logs
```
- Work as normal, i.e. edit files, create new files, delete files. 
- You now *stage* your changes, e.g.
```{r, engine = 'bash', eval = FALSE}
      git add foo.file.1 foo.file.2 # for edits or new files
      git rm foo.file.3 # for removed files
```
- You then *commit* these:
```{r, engine = 'bash', eval = FALSE}
      git commit # Editor opens for your log msg
```


# Git: Tiny tutorial (2)

- Pull down any recent changes by others from the remote:
```{r, engine = 'bash', eval = FALSE}
      git pull
      git log # If new changes, read their logs
```
- Now, push your own changes to the remote
```{r, engine = 'bash', eval = FALSE}
      git push # requires permissions
```

- Undo changes:
```{r, engine = 'bash', eval = FALSE}
      git reset a381f2f # move back "head"
      git revert a381f2f # applies new change to revert 
```


# References
